Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-30 Manuscript/manuscript.Rmd
intro edits

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].

## Exhisting problems and limitations

A growing number of studies have proposed computational tools used to identify additional key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions. However, the growing number of publically available annotated genomic elements, and there collinearity, pose a significant problem for establishing methods robust enough to determine the most important genomic factors that contribute to TAD boudandary formation. Likewise, it is unclear what the best method is for quantifying this relationship. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

Another key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. It is unclear how varying levels of resolution affect the association between functional genomic annotations and TAD boundary regions. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing].

## Previous methods

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites 
on GM12878 [@mourad2016computational].

## PROBLEMS AND LIMITATIONS

The distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches toward TAD boundary prediction  <!--[@dixon2015chromatin; @di2017novo]-->. Currently, classification models have been used to draw associations between these functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting the performance of predictive modeling.

First, the use of different levels of association between TAD boundary regions and genomic features as predictors remain underexplored. These associations are termed predictor-types and make up the feature space of classifiers. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association as it simply counts the total number of peaks defining a genomic annotation that are inside a given TAD boundary region. Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of overlap between a genomic annotation and a boundary region. However, the distance between boundary regions and genomic annotations are currently unexplored. This association allows for quantifying the spatial relationship between genomic elements and boundary regions. We hypothesize that the inclusion of this type of association will correspond with more accurate identification of functional genomic elements association with TAD boundary formation.

Secondly, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These boundary regions make up the minority class, while the non-boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *class imbalance* (CI). Here, CI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. That is, CI is bounded by 0 and 0.50. A dataset is considered balanced if CI $\approx$ 0.50, and imbalanced as CI approaches 0. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. We hypothesize that the implementation of novel data resampling techniques will significantly improve model performance. Futhermore, the choice of performance metric can mask the poor performance of models implemented on imbalanced data.  It has been shown that conventional metrics such as AUC and accuracy are insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. Therefore, we hypothesize that a more composite metric such as Matthews Correlation Coefficient (MCC), might provide more sensitive assessment of model performance, given that it penalizes for both false positives and false negatives.

Thirdly, TAD boundary location is known to be heavily influenced by the resolution of the subsequent Hi-C interaction matrix [@sexton2012three]. Higher resolutions require a larger number of mapped reads, thereby creating larger data matrices, increasing computation time. However, the increased resolution available in newer datasets has highlighted the existence of a hierarchical structure of TADs inside other TADs [@pal2019hi]. It is unclear how varying levels of resolution affect the association between functional genomic annotations and TAD boundary regions.
<!--[link these two paragraphs]-->
<!--[Paragraph about disconnect between Hi-C and epigenomic data] 
Despite the success of Hi-C technologies in revealing the high-order levels of genome organization (chromosome territories, A/B compartments, TADs, sub-TADs, chromatin loops), resolution of Hi-C data (~750kb [??? REF, Bonev 2017?]) remains well below resolution of epigenomic assays (hundredths of bases for ChIP-seq). One of the key missing pieces to connect 3D genome architecture to transcription has been the inability of Hi-C to attain the resolution necessary to discern gene-level features of chromatin. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation of gene expression and its potential link to chromatin architecture. Although computational [??? REFS?](Zhang et al., “HiCPlus.”, Carron et al., “Boost-HiC.”) and technological [@Hsieh:2019aa][@Ma:1998aa][@Belaghzal2017] methods for enhancing resolution of Hi-C data started to appear, they [??? remains to write about their pitfalls]. One aim of this study is to predict the location of TAD boundaries on a scale comparable to epigenomic assays. For this, we developed a model trained on a relatively coarse Hi-C data for predicting 3D features (TAD boundaries) from 1D data (epigenomic datasets).-->

## PREVIOUS METHODS

<!--[still revising]-->
Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites 
on GM12878 [@mourad2016computational]. Based on a 1 kb binned genomic data matrix 
and 20 kb flanked TAD boundaries, the model was trained using proportions of 
overlaps between ChiP-seq peak defined functional element coordinates and genomic 
bins. However, several problems exist with this study. First, many transcription 
factors are known to colocalize, and therefore exhibit high correlations. Given 
a group of features that are highly correlated, L1-regularization tends to pick 
only one of them, while shrinking the others to zero [@zou2005regularization]. 
Here, ensemble learning techniques could offer a more robust method toward 
identifying  annotations associated with TAD boudaries. Likewise, it is unclear 
how other sets of epigenomic features affect prediction, including chromatin state 
segmentations and histone modifications. Secondly, the class imbalance issue was 
not addressed prior to modelling. Instead, model performance was evaluated using 
ROC (receiver operating characteristic) curves and subsequent AUC (area under the 
curve). It has been shown that AUC is insensitive to class imbalance, creating 
artificially inflated values inlfuenced by the majority class [@prati2011survey; 
@wei2013role]. It is unclear how simple data level solutions to class imbalance, 
such random under-sampling and random over-sampling affect TAD boundary prediction. 
Likewise, a more composite metric such as Matthews Correlation Coefficient (MCC), 
might provide more sensitive assessment of model performance. Thirdly, the relationship 
between annotations and boundaries was quantified using proportions of overlaps. This 
type of predictor ignores distances, which offers a much more spatially descriptive 
feature space and could improve predictive performance. Lastly, it is unclear how *HiCfeat* 
would perform on other cell lines and Hi-C data resolutions, giving *HiCfeat* a limited 
application in TAD boundary prediction.

Another method used an L2-based regularized linear model to discriminate between what was termed as "boundary-" and "TAD-segments" of the hESC cell line [@hong2017computational].

A method devolped in 2015 used a tree based model to predict TAD boundary regions using 9 different cell line specific histone modifications, as well as consensus CTCF peaks, across 8 cell lines [@huang2015predicting].

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed the spatial organization of the human genome  [@lieberman2009comprehensive]. The genome has been shown to form tight, compact folds in order to fit inside the nucleus. This folding does not occur randomly; in fact, the genome exhibits a hierarchical 3D structure [@dixon2012topological]. Among this hierarchy are distict compartments where the genome folds and interacts with itself more frequently termed topologically associating domains (TADs). TADs have garnered much interest in recent years given that they are heavily linked to gene expression [@dixon2015chromatin]. Moreover, changing the 3D structure of the DNA, causing disruptions in the boundaries that demarcate TADs, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. It has been thought that the collection of particular functional genomic elements may contribute to the formation of TADs. Namely, the CCCTC-binding factor (CTCF) has been shown to be enriched at TAD boundaries [@ghirlando2016ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, chromatin has been shown to exhibit certain modifications at TAD boundaries, evidenced by the frequent observation of the H3K36me3 histone mark [@dixon2012topological].

A growing number of studies have proposed computational tools used to identify the complete set of key molecular drivers of the formation of TAD boundaries. The classical approach is to build classification models that leverage the relationship between ChIP-seq genomic annotations and TAD boundary regions. A key factor in these approaches becomes resolution. Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Due to high sequencing cost, most available Hi-C datasets have relatively low resolution, in the kilobase range. This is in stark contrast to ChIP-seq experiments which annotate regions of the genome at hundreds of base pairs in resolution [??? REF, Bonev 2017?]. This resolution gap between 1D and 3D genomic maps, as well as a paucity of chromatin features below the level of TADs, has significantly limited our understanding of epigenomic regulation its potential link to chromatin architecture. However, simply increasing sequencing depth to obtain higher resolution could have unintended consequences since this could introduce sparsity in the number of contacts which TAD callers are often ill-equiped to handle [@cresswell2019spectraltad, @sexton2012three]. Increased resolution could also further exacerbate the class imbalance created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. It is unclear how simple data level solutions to class imbalance, such random under-sampling and random over-sampling affect TAD boundary prediction. Additionally, how relationships between genomic anntotations and bins of varying resolutions are quantified remains an open question. Incorporating a distance measure has largely been ignored, but could offer a much more spatially descriptive feature space thereby improving predictive performance.

Recently, an L1-regularized multiple logistic regression model, called *HiCfeat*, 
was developed to predict TAD boundaries using transcription factor binding sites (TFBS) 
on GM12878 [@mourad2016computational]. High resolution Hi-C data at 1 kb was used and percentage of overlap between regions defining TFBS and 1 kb bins was used for feature engineering. However, class imbalance was not adressed and, instead area under the receiver operating curve (AUROC) was used to evaluate model performance. This is problematic since it has been shown that AUC is insensitive to class imbalance, creating artificially inflated values inlfuenced by the majority class [@prati2011survey; @wei2013role]. Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase hypersensitive sites, and histone modifications in an L2-based regularized linear model [@hong2017computational]. Both used re-sampling techniques prior to modeling, with both implementing random under-sampling to balance TAD and non-TAD boundary regions. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Furthermore, there was not a consensus in either resolution or feature engineering. The BART model was implemented on 5 kb Hi-C data using sequence read count between ChIP-seq data and 5 kb bins, while the L2-based regularized linear model was implemented on filtered TADs greater than 200 kb in length using the elemental count that appeared in each 150 kb bin.
